---
title: "Relational Data - Chap 10"
author: "Mac Faldet"
date: "November 9, 2017"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
library(tidyverse)
library(readr)
library(dplyr)
library(nycflights13)
```

## Chapter 13 Relational Data

For nycflights13:

`flights` connects to planes via a single variable, `tailnum`.

`flights` connects to airlines through the `carrier` variable.

`flights` connects to airports in two ways: via the `origin` and `dest` variables.

`flights` connects to weather via `origin` (the location), and `year`, `month`, `day` and `hour` (the time).


Topics, functions:

  keys: `primary key`, `foreign key`,
  mutating joins: `left_join`, `right_join`, `inner_join`, `full_join`
merge vs. joins
  filtering joins: `semi_join`, `anti_join`
  set operations: `intersect`, `union`, `setdiff`


## 13.2.1 Exercises

1. Imagine you wanted to draw (approximately) the route each plane flies from its origin to its destination. What variables would you need? What tables would you need to combine?

  - We would need the `origin`, `dest`, and `carrier` variables in order to join on the *airplanes* and *airports* tibbles.

2. I forgot to draw the relationship between weather and airports. What is the relationship and how should it appear in the diagram?

  - you can match the variable `origin` in *weather* to `faa` in *airports*.

3. weather only contains information for the origin (NYC) airports. If it contained weather records for all airports in the USA, what additional relation would it define with flights?

  - `year`, `month`, `day`, `hour`, and `origin` would all additionally match.

4. We know that some days of the year are “special”, and fewer people than usual fly on them. How might you represent that data as a data frame? What would be the primary keys of that table? How would it connect to the existing tables?

  - I would create a tibble of special time frames in which a certain flighing behavior was expected to happen, then I would have it join to *flight* on `year`, `month`, `day`, and `hour` if needed.
  
  
## 13.3 Keys

There are two types of keys:

  - A _*primary key*_ uniquely identifies an observation in its own table. For example, planes$tailnum is a primary key because it uniquely identifies each plane in the planes table.

  - A _*foreign key*_ uniquely identifies an observation in another table. For example, the flights$tailnum is a foreign key because it appears in the flights table where it matches each flight to a unique plane.


## 13.3.1 Exercises

1. Add a surrogate key to flights.

```{r surrogate, echo=FALSE}
flights %>% 
  mutate(flight_id = row_number())
```

2. Identify the keys in the following datasets

  1. Lahman::Batting,
  2. babynames::babynames
  3. nasaweather::atmos
  4. fueleconomy::vehicles
  5. ggplot2::diamonds
(You might need to install some packages and read some documentation.)

  - The keys are grouped together below in a group by, proving themselves

``{r keys, echo=FALSE}
Lahman::Batting %>%
  group_by(playerID, yearID, stint) %>%
  filter(n() > 1) %>%
  nrow()

babynames::babynames %>%
  group_by(year, sex, name) %>%
  filter(n() > 1) %>%
  nrow()

nasaweather::atmos %>%
  group_by(lat, long, year, month) %>%
  filter(n() > 1) %>%
  nrow()

fueleconomy::vehicles %>%
  group_by(id) %>%
  filter(n() > 1) %>%
  nrow()

ggplot2::diamonds %>%
  distinct() %>% 
  nrow()
``

  - Note in diamonds: even using all variables there is still thousands of rows, there is no identifying variable.

3. Draw a diagram illustrating the connections between the Batting, Master, and Salaries tables in the Lahman package. Draw another diagram that shows the relationship between Master, Managers, AwardsManagers.

How would you characterise the relationship between the Batting, Pitching, and Fielding tables?

  - I would characterism them as very similar, given they are the only tables with a unique primary `stint` identifier that is unique for each player across leagues.
  
## 13.4.6 Exercises

1.  Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:

```{r foursixone,echo=TRUE}
airports %>%
  semi_join(flights, c("faa" = "dest")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```
(Don’t worry if you don’t understand what semi_join() does — you’ll learn about it next.)

You might want to use the size or colour of the points to display the average delay for each airport.

```{r location,echo=FALSE}
avg_dest_delays <-
  flights %>%
  group_by(dest) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c(dest = "faa")) %>%
  ggplot(aes(lon, lat, colour = delay, size = delay)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

2.  Add the location of the origin and destination (i.e. the lat and lon) to flights.

```{r latlon,echo=FALSE}
flights %>%
  left_join(airports, by = c(dest = "faa")) %>%
  left_join(airports, by = c(origin = "faa"))
```

3.  Is there a relationship between the age of a plane and its delays?

  - No, if anything delays perhaps improve over their age but are never easily predictable based off its age.

```{r agedelay,echo=FALSE}
plane_ages <- 
  planes %>%
  mutate(age = 2013 - year) %>%
  select(tailnum, age)

flights %>%
  inner_join(plane_ages, by = "tailnum") %>%
  group_by(age) %>%
  filter(!is.na(dep_delay)) %>%
  summarise(delay = mean(dep_delay)) %>%
  ggplot(aes(x = age, y = delay)) +
  geom_point() +
  geom_line()
```

4.  What weather conditions make it more likely to see a delay?

  - Any sort of rain causes a delay. With no rain there is the fewest delays.

```{r precipitation,echo=FALSE}
flights %>%
  inner_join(weather, by = c("origin" = "origin",
                            "year" = "year",
                            "month" = "month",
                            "day" = "day",
                            "hour" = "hour")) %>%
  group_by(precip) %>%
  summarise(delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = precip, y = delay)) +
    geom_line() + geom_point()
```

5.  What happened on June 13 2013? Display the spatial pattern of delays, and then use Google to cross-reference with the weather.

  - Wiki says there were loads of storms everywhere.
  
```{r wikicheck,echo=FALSE}
flights %>%
  filter(year == 2013, month == 6, day == 13) %>%
  group_by(dest) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c("dest" = "faa")) %>%
  ggplot(aes(y = lat, x = lon, size = delay, colour = delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap()
```

### 13.4 Other Implications

base::merge() can perform all four types of mutating join:

  dplyr	                merge
inner_join(x, y)    	merge(x, y)
left_join(x, y)     	merge(x, y, all.x = TRUE)
right_join(x, y)    	merge(x, y, all.y = TRUE),
full_join(x, y)     	merge(x, y, all.x = TRUE, all.y = TRUE)

SQL is the inspiration for dplyr’s conventions, so the translation is straightforward:

  dplyr	                          SQL
inner_join(x, y, by = "z")	    SELECT * FROM x INNER JOIN y USING (z)
left_join(x, y, by = "z")	      SELECT * FROM x LEFT OUTER JOIN y USING (z)
right_join(x, y, by = "z")    	SELECT * FROM x RIGHT OUTER JOIN y USING (z)
full_join(x, y, by = "z")	      SELECT * FROM x FULL OUTER JOIN y USING (z)
  - Note that “INNER” and “OUTER” are optional, and often omitted.

Joining different variables between the tables, e.g. inner_join(x, y, by = c("a" = "b")) uses a slightly different syntax in SQL: SELECT * FROM x INNER JOIN y ON x.a = y.b. As this syntax suggests, SQL supports a wider range of join types than dplyr because you can connect the tables using constraints other than equality (sometimes called non-equijoins).

## 13.5 Filtering Joins

Filtering joins match observations in the same way as mutating joins, but affect the observations, not the variables. There are two types:

  - `semi_join(x, y)` _*keeps all*_ observations in `x` that have a match in `y`.
  - `anti_join(x, y)` _*drops all*_ observations in `x` that have a match in `y`.

Semi-joins are useful for matching filtered summary tables back to the original rows. For example, imagine you’ve found the top ten most popular destinations:

```{r example,echo=TRUE}
top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)
top_dest

flights %>% 
  semi_join(top_dest)
```

How many flights are not found in planes

```{r antiexample,echo=TRUE}
flights %>%
  anti_join(planes, by = "tailnum") %>%
  count(tailnum, sort = TRUE)
```

## 13.5.1 Exercises

1. What does it mean for a flight to have a missing `tailnum`? What do the tail numbers that don’t have a matching record in `planes` have in common? (Hint: one variable explains ~90% of the problems.)

  - The carriers AA and MQ don't report tailnumbers

2. Filter flights to only show flights with planes that have flown at least 100 flights.

```{r hunnidflights,echo=FALSE}
hunnidplanes <- 
  filter(flights) %>%
  group_by(tailnum) %>%
  count() %>%
  filter(n > 100)

flights %>%
  semi_join(hunnidplanes, by = "tailnum")
```

3. Combine `fueleconomy::vehicles` and `fueleconomy::common` to find only the records for the most common models.

```{r combine,echo=FALSE}
glimpse(fueleconomy::vehicles)
glimpse(fueleconomy::common)
fueleconomy::vehicles %>%
  semi_join(fueleconomy::common, by = c("make", "model"))
```

4. Find the 48 hours (over the course of the whole year) that have the worst delays. Cross-reference it with the `weather` data. Can you see any patterns?

  - June 13, 2013

5. What does `anti_join(flights, airports, by = c("dest" = "faa"))` tell you? What does `anti_join(airports, flights, by = c("faa" = "dest"))` tell you?

  - `anti_join(flights, airports, by = c("dest" = "faa"))` gives all airports that do NOT have a destination in airports.
  
  - `anti_join(airports, flights, by = c("faa" = "dest"))` gives all the airports that didn't have plane data at them.

6. You might expect that there’s an implicit relationship between plane and airline, because each plane is flown by a single airline. Confirm or reject this hypothesis using the tools you’ve learned above.

  - I reject this! I'm pretty sure planes can be used by others, or sold at another airline, but i shall check.
  
```{r airlinecheck,echo=FALSE}
flights %>%
  group_by(tailnum, carrier) %>%
  count() %>%
  filter(n() > 1) %>%
  select(tailnum) %>%
  distinct()
```


